Data Specialization for a Staged Lambda Calculus

### 2015-02-02 paper/semantics.tex
small fixes and edits

The input to our splitting algorithm is given in \lang, a typed, two-stage
lambda calculus with products, sums, and isorecursive types. This language is
adapted from \cite{davies96}, and is very similar to the staged languages
considered in the partial evaluation and metaprogramming literature. In this
paper, we restrict our attention to a two-stage language to simplify our
presentation of the splitting algorithm, but we expect that the techniques
presented here extend to more stages.

Our dynamic semantics of \lang\ differ from those given in \cite{davies96} and
the metaprogramming literature, because we avoid duplicating stage \bbtwo\
computations. In addition to being more efficient, our non-duplicating semantics
are easier to compare with our splitting algorithm, whose development and
exposition is our main goal.

@@

In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
stage~\bbone\ term. To ensure that the staging features are only applied to
valid terms, we index our judgments not only by a type, but also a stage $w$.

The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
the context $\Gamma$, and is defined in \cref{fig:statics}.

>>>>

The input to our splitting algorithm is given in \lang, a typed, modal
lambda calculus with products, sums, and isorecursive types. This language is
adapted from \cite{davies96}, and is very similar to the staged languages
considered in the partial evaluation and metaprogramming literature,
though it differs in three key ways:

\begin{enumerate}

\item We restrict our attention to a two-stage language to simplify our
presentation of the splitting algorithm, though we expect that the techniques
presented here extend to more stages.

\item Although our language describes computations with two {\em stages}
(\bbone\ and \bbtwo), its type system makes use of three {\em worlds}
(\bbonem, \bbonep, \bbtwo).  For reasons discussed later, we require
that stage \bbone\ computations be classified as being either pure
(occuring at \bbonep) or mixed (occuring at \bbonem). 
Only mixed code may transition to stage/world \bbtwo.

\item Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
the metaprogramming literature, because we avoid duplicating stage \bbtwo\
computations. In addition to being more efficient, our non-duplicating semantics
are easier to compare with our splitting algorithm, whose development and
exposition is our main goal.

\end{enumerate}

@@

To ensure that the staging features are only applied in a valid way, 
we index our judgments not only by a type, but also a stage $w$.
The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
the context $\Gamma$, and is defined in \cref{fig:statics}.
